An Extendable \proglang{Python} Implementation of Robust Optimisation Monte Carlo

### 2022-02-20 article/latex_files/ch1/1-introduction.tex

Simulator-based models are particularly captivating due to the
modelling freedom they provide. In essence, a simulator-based model can
describe any data generating mechanism that can be written as a finite
set of algorithmic steps. This modelling freedom comes at a cost;
performing the inference, i.e., sample or evaluate the posterior
distribution, is challenging.

Optimization Monte Carlo (OMC) proposed by~\citet{Meeds2015} was a
novel LFI approach for approximating the posterior
distribution. Afterwards, \citet{Forneron2016} provided a similar
method naming it `reverse sampler'. The central idea is turning the
stochastic data-generating mechanism into a set of
fully-parallelizable deterministic optimisation processes. In their
work, ~\citet{Ikonomov2019}, located some critical failure modes of
OMC, and proposed Robust OMC (ROMC) an alternative version of OMC with
the appropriate improvements.

In this paper, we present an extendible and parallelisable
implementation of ROMC at the \proglang{Python} package \pkg{Engine
  for Likelihood-Free inference (ELFI)}. We choose carefully the
designing principles for making the implementation extendible. As we
describe analytically in the next chapter, ROMC can be understood not
only as a specific algorithm, but mainly as a framework for obtaining
weighted samples from the posterior; it defines a sequence of
algorithmic steps, without enforcing a specific algorithm for solving
them. Therefore, a researcher may adopt ROMC's backbone approach, but
experiment with alternative methods to solve the tasks. We have
designed ROMC for facilitating such alterations. Finally, we have
tested the accuracy and the efficiency of our implementation on the
LFI examples supported by the \pkg{ELFI} package.

To the best of our knowledge, this is the first attempt to implement
the ROMC inference method to a generic LFI framework. Therefore, there
are not other implementation directly comparable to ours. For this
reason, we test the implementation against (i) an artificial example
with tractable likelihood and (ii) the second-order moving average
(MA2) example from the \pkg{ELFI} package. In the latter, we generate
data using known parameters and we compare our results with the
typical Rejection ABC approach.

>>>>

Simulator-based models are particularly captivating due to the
modelling freedom they provide. In essence, a simulator-based model
can describe any data generating mechanism that can be written as a
finite set of algorithmic steps. Such freedom comes at a cost;
performing the inference, i.e., sampling or evaluating the posterior,
is challenging.

Optimisation Monte Carlo (OMC) proposed by~\citet{Meeds2015} is a
novel LFI approach for approximating the posterior distribution. The
central idea is turning the stochastic data-generating mechanism into
a set of deterministic optimisation processes. Afterwards,
\citet{Forneron2016} provided a similar method under the name `reverse
sampler'. In their work, ~\citet{Ikonomov2019}, located some critical
limitations of OMC. They proposed Robust OMC (ROMC), an alternative
version of OMC with appropriate modifications.

In this paper, we present an implementation of ROMC at the
\proglang{Python} package \pkg{Engine for Likelihood-Free inference
(ELFI)}. We follow appropriate design principles for ensuring
extensibility. As we describe analytically in the next chapter, ROMC
is a general framework for obtaining weighted samples from the
posterior; it defines a sequence of algorithmic steps without
enforcing a specific algorithm for solving each step\footnote{For
being a ready-to-use algorithm,~\citet{Ikonomov2019} proposed a
default method for each step, but this choice is by no means
restrictive.}. A researcher may adopt ROMC and develop novel methods
to solve each task. We have designed our software for facilitating
such alterations. Finally, we have tested the accuracy and the
efficiency of our implementation on some LFI examples supported by the
\pkg{ELFI} package.

To the best of our knowledge, this is the first attempt to implement
the ROMC inference method to a generic LFI framework; there are no
other packages for making direct comparisons. Therefore, we test it
against (i) an artificial example with tractable likelihood and (ii)
the second-order moving average (MA2) example from the \pkg{ELFI}
package. In the latter, we generate data using known parameters and
compare our results with the typical Rejection ABC~\citet{Beaumont2002}.

### 2022-02-20 article/latex_files/ch2/2-1-implicit-models.tex

An implicit or simulator-based model is a parameterised stochastic
data generating mechanism. The key characteristic of these models is
that we can sample data points, but we cannot evaluate the
likelihood. Formally, a simulator-based model is a parameterised
family of probability density functions
\(\{ p(\yb|\thetab) \}_{\thetab}\) whose closed-form is either unknown
or computationally intractable. Practically, a simulator
\( M_r(\thetab) \) can be understood as a random black-box machine
(computer code) that generates samples \(\yb\) in a stochastic manner
for any given a set of parameters \(\thetab\), i.e.
\( M_r(\thetab) \rightarrow \yb \). We can isolate the randomness by
introducing the stochastic nuisance variables \(\vb \sim p(\vb)\) so
that the simulator becomes a deterministic mapping \(\simulator\) that
maps \((\thetab, \vb)\) to the data \(\yb\), i.e.\
\(\yb=\simulator(\thetab,\vb)\).  In terms of computer code,
\(p(\vb)\) is defined as a random number generating process within the
computer code.  % Importantly, we can fix the value of \(\vb\) while
% changing \(\thetab\) by fixing the random seed used.

@@

The modelling freedom of simulator-based models, however, comes at the
price of difficulties in inferring their parameters. Denoting observed
data by \(\data\), the main difficulty is that the likelihood function
\(L(\thetab) = p(\data|\thetab)\) is generally intractable. To better
see the sources of the intractability, and to address them, we go back
to the basic characterisation of the likelihood as the (rescaled)
probability of a parameter of the model to generate data \(\yb\) that
is similar to the observed data \(\data\). More formally, the
likelihood \(L(\thetab)\) equals
\begin{equation} \label{eq:likelihood}
  L(\thetab) = \lim_{\epsilon \to 0} c_\epsilon \int_{\yb \in B_{d,\epsilon}(\data)} p(\yb|\thetab)d\yb =
  \lim_{\epsilon \to 0} c_\epsilon \Pr(\simulator(\thetab, \vb) \in \region(\data)  \mid \theta)
\end{equation}
where \(c_\epsilon\) is a proportionality factor that depends on
\(\epsilon\) and \(\region(\data)\) is an \(\epsilon\) region around \(\data\)
that is defined via a distance function \(d\), i.e.\ \(\region(\data)
:= \{\yb: d(\yb, \data) \leq \epsilon \}\).

The basic characterisation of the likelihood in \eqref{eq:likelihood}
highlights two sources of intractability: The first is the computation
of the probability
\(\Pr(\simulator(\thetab,\vb) \in \region(\data))\), the second is the
limit of \(\epsilon \to 0\).  Approximating the probability with
samples becomes computationally infeasible if \(\epsilon\) is too
small. Hence, a large class of inference methods work with
\(\epsilon > 0\), which leads to the approximate likelihood function
\(L_{d, \epsilon}(\thetab)\),
\begin{equation} \label{eq:approx_likelihood}
  L_{d, \epsilon}(\thetab) = \Pr(\yb \in \region(\data) \mid \theta), \quad \text{where  } \epsilon > 0.
\end{equation}

and, in turn, to the approximate posterior,

>>>>

An implicit or simulator-based model is a parameterised stochastic
data generating mechanism. The key characteristic of these models is
that we can sample data points, but we cannot evaluate the
likelihood. Formally, a simulator-based model is a parameterised
family of probability density functions
\(\{ p(\yb|\thetab) \}_{\thetab}\) whose closed-form is either unknown
or computationally intractable. Practically, in these scenarios, we
can only a access a simulator \( M_r(\thetab) \), i.e. a random
black-box machine (computer code) that generates samples \(\yb\) in a
stochastic manner for any given a set of parameters \(\thetab\);
\( M_r(\thetab) \rightarrow \yb \). We can isolate the randomness by
introducing the stochastic nuisance variables \(\ub \sim p(\ub)\) so
that the simulator becomes a deterministic mapping \(\simulator\) that
maps \((\thetab, \ub)\) to the data \(\yb\), i.e.\
\(\yb=\simulator(\thetab,\ub)\). Within the computer code, the
distribution \(p(\ub)\) is defined as the random number generating
process.

@@

The modelling freedom of simulator-based models, however, comes at the
price of difficulties in inferring their parameters. Denoting the
observed data as \(\data\), the main difficulty is that the likelihood
function \(L(\thetab) = p(\data|\thetab)\) is generally
intractable. To better see the sources of the intractability, and to
address them, we go back to the basic characterisation of the
likelihood as the (rescaled) probability of a parameter of the model
to generate data \(\yb\) that is similar to the observed data
\(\data\). More formally, the likelihood \(L(\thetab)\) equals
\begin{equation} \label{eq:likelihood}
  L(\thetab) = \lim_{\epsilon \to 0} c_\epsilon \int_{\yb \in B_{d,\epsilon}(\data)} p(\yb|\thetab)d\yb =
  \lim_{\epsilon \to 0} c_\epsilon \Pr(\simulator(\thetab, \ub) \in \region(\data)  \mid \theta)
\end{equation}
where \(c_\epsilon\) is a proportionality factor that depends on
\(\epsilon\) and \(\region(\data)\) is an \(\epsilon\) region around \(\data\)
that is defined via a distance function \(d\), i.e.\ \(\region(\data)
:= \{\yb: d(\yb, \data) \leq \epsilon \}\).

The basic characterisation of the likelihood in \eqref{eq:likelihood}
highlights two sources of intractability: The first is the computation
of the probability
\(\Pr(\simulator(\thetab,\ub) \in \region(\data))\), the second is the
limit of \(\epsilon \to 0\).  Approximating the probability with
samples becomes computationally infeasible if \(\epsilon\) is too
small. Hence, a large class of inference methods work with
\(\epsilon > 0\), which leads to the approximate likelihood function
\(L_{d, \epsilon}(\thetab)\)
\begin{equation} \label{eq:approx_likelihood}
  L_{d, \epsilon}(\thetab) = \Pr(\yb \in \region(\data) \mid \theta), \quad \text{where  } \epsilon > 0.
\end{equation}

and, in turn, to the approximate posterior

### 2022-02-20 article/latex_files/ch4/4-2-inference.tex

\noindent
This is the basic inference utility of the ROMC implementation; we
draw \(n_2\) samples for each bounding box region. This gives a total of
\(k \times n_2\), where \(k < n_1\) is the number of the optimal points
remained after filtering\footnote{From the \(n_1\) optimisation
  problems, only the ones with \(d_i(\thetab_*) < \epsilon\) are
  maintained for building a bounding box}. The samples are drawn from
a uniform distribution \(q_i\) defined over the corresponding bounding
box and the weight \(w_i\) is computed as in
Algorithm~\ref{alg:sampling_GB}.

@@

\noindent
The function \code{compute_expectation} computes the expectation
\(\E_{p(\thetab|\data)}[h(\thetab)]\) using
expression~\eqref{eq:expectation}. The argument \code{h} can be any
python \code{Callable}. Method \code{eval_unnorm_posterior} computes
the unnormalised posterior approximation using
expression~\eqref{eq:approx_posterior}. Method \code{eval_posterior}
evaluates the normalised posterior estimating the partition function
\(Z = \int p_{d,\epsilon}(\thetab|\data)d\thetab\); with Riemann's
integral approximation. The approximation is computationally feasible
only in low-dimensional parametric spaces.

>>>>

\noindent
This is the basic inference utility of the ROMC implementation; we
draw \(n_2\) samples for each bounding box region. This gives a total of
\(k \times n_2\), where \(k < n_1\) is the number of the optimal points
remained after filtering\footnote{From the \(n_1\) optimisation
  problems, only the ones with \(d_i(\thetab^*) < \epsilon\) are
  maintained for building a bounding box}. The samples are drawn from
a uniform distribution \(q_i\) defined over the corresponding bounding
box and the weight \(w_i\) is computed as in
Algorithm~\ref{alg:sampling_GB}.

@@

\noindent
The function \code{compute_expectation} computes the expectation
\(\E_{p(\thetab|\data)}[h(\thetab)]\) using
expression~\eqref{eq:expectation}. The argument \code{h} can be any
python \code{Callable}. The method \code{eval_unnorm_posterior} computes
the unnormalised posterior approximation using the
expression~\eqref{eq:approx_posterior}. The method \code{eval_posterior}
evaluates the normalised posterior estimating the partition function
\(Z = \int p_{d,\epsilon}(\thetab|\data)d\thetab\); with Riemann's
integral approximation. The approximation is computationally feasible
only in a low-dimensional parametric space.

### 2022-02-20 article/latex_files/ch5/5-ma2.tex
small change

% \begin{figure}[ht]
%     \begin{center}
%       \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter4/mae2_prior_samples.png}
%     \end{center}
%     \caption[MA2 example, prior distribution.]{Prior distribution proposed by \cite{Marin2012}. The
%       samples follow a triangular shape.}
%   \label{fig:ma2_1}
% \end{figure}

@@

\begin{table}
\begin{center}
\begin{tabular}{ c|c|c|c|c }
\hline
& \(\mu_{\theta_1}\) & \(\sigma_{\theta_1}\) & \(\mu_{\theta_2}\) & \(\sigma_{\theta_2}\) \\
\hline \hline
Rejection ABC & 0.516 & 0.142 & 0.07 & 0.172 \\
\hline
ROMC (gradient-based) & 0.503 & 0.143 & 0.032 & 0.17 \\
\hline
ROMC (Bayesian optimisation) & 0.494 & 0.16 & 0.086 & 0.167 \\
\hline
ROMC (Neural Network) & 0.503 & 0.140 & 0.03 & 0.171 \\
\hline
\end{tabular}
\end{center}
\caption{Comparison of the samples obtained from the estimated
  posterior with (a) Rejection sampling and (b) the different versions
  of ROMC. We observe that the obtained samples share similar
  statistics along all methods. \label{tab:ma2}}
\end{table}

\begin{figure}[ht]
    \begin{center}
        % \input{./latex_files/images/chapter4/ma2_region_3.tex}
        % \input{./latex_files/images/chapter4/ma2_region_3_bo.tex}
        \includegraphics[width=0.49\textwidth]{./latex_files/images/chapter4/ma2_region_1.png}
        \includegraphics[width=0.49\textwidth]{./latex_files/images/chapter4/ma2_region_1_bo.png}
    \end{center}
  \caption[The acceptance region of a specific deterministic simulator.]{The acceptance region in a specific optimisation problem. In the left figure the region obtained with gradient-based optimiser and in the right one with Bayesian Optimisation.}
  \label{fig:ma2_5}
\end{figure}

@@

As stated above, ROMC is a ridiculously parallelisable
method. Therefore, it is straightforward to parallelise the fitting
part, i.e. (i) solving the optimisation problems and (ii) estimating
the proposal regions. Our implementation supports exploiting all the
available CPU cores through the built-in \proglang{Python} package
\pkg{multiprocess}\footnote{https://docs.python.org/3/library/multiprocessing.html}. In
Figure~\ref{fig:exec_parallel} we observe the execution times for
performing the inference on the ma2 model; the parallel version
performs both tasks almost six times faster than the sequential. The
result is reasonable given that the experiments have run in a single
machine with the Intel® Core™ i7-8750H Processor, which has six
separate cores.

\begin{figure}[ht]
  \begin{center}
    \resizebox{.49\columnwidth}{!}{%
      \input{./latex_files/images/chapter4/exec_time_solve.tex}
    }
    \resizebox{.49\columnwidth}{!}{%
      \input{./latex_files/images/chapter4/exec_time_regions.tex}
    }
    \end{center}
    \caption[Execution time exploiting parallelisation]{Comparison
      between parallel and sequential execution of the fitting part of
      ROMC. We observe that the parallel version runs between 2.5 an 6
      times faster.}
      \label{fig:exec_parallel}
\end{figure}

>>>>

@@

\begin{table}
\begin{center}
\begin{tabular}{ c|c|c|c|c }
\hline
& \(\mu_{\theta_1}\) & \(\sigma_{\theta_1}\) & \(\mu_{\theta_2}\) & \(\sigma_{\theta_2}\) \\
\hline \hline
Rejection ABC & 0.516 & 0.142 & 0.07 & 0.172 \\
\hline
ROMC (gradient-based) & 0.501 & 0.142 & 0.033 & 0.169 \\
\hline
ROMC (Bayesian optimisation) & 0.494 & 0.16 & 0.086 & 0.167 \\
\hline
ROMC (Neural Network) & 0.491 & 0.138 & 0.04 & 0.172 \\
\hline
\end{tabular}
\end{center}
\caption{Comparison of the samples obtained from the estimated
  posterior with (a) Rejection sampling and (b) the different versions
  of ROMC. We observe that the obtained samples share similar
  statistics along all methods. \label{tab:ma2}}
\end{table}

\begin{figure}[ht]
    \begin{center}
        \includegraphics[width=0.49\textwidth]{./latex_files/images/chapter4/ma2_region_1.png}
        \includegraphics[width=0.49\textwidth]{./latex_files/images/chapter4/ma2_region_1_bo.png}
    \end{center}
  \caption[The acceptance region of a specific deterministic simulator.]{The acceptance region in a specific optimisation problem. In the left figure the region obtained with gradient-based optimiser and in the right one with Bayesian Optimisation.}
  \label{fig:ma2_5}
\end{figure}

@@

As stated above, ROMC is a ridiculously parallelisable
method. Therefore, it is straightforward to parallelise the fitting
part, i.e. (i) solving the optimisation problems and (ii) estimating
the proposal regions. Our implementation supports exploiting all the
available CPU cores through the built-in \proglang{Python} package
\pkg{multiprocess}\footnote{https://docs.python.org/3/library/multiprocessing.html}. In
Figure~\ref{fig:exec_parallel} we observe the execution times for
performing the inference on the MA2 model; the parallel version
performs both tasks almost five times faster than the sequential. The
result is reasonable given that the experiments have run in a single
machine with the Intel® Core™ i7-8750H Processor, which has six
separate cores.

\begin{figure}[ht]
  \begin{center}
    \resizebox{.49\columnwidth}{!}{%
      \input{./latex_files/images/chapter4/exec_time_solve.tex}
    }
    \resizebox{.49\columnwidth}{!}{%
      \input{./latex_files/images/chapter4/exec_time_regions.tex}
    }
    \end{center}
    \caption[Execution time exploiting parallelisation]{Comparison
      between parallel and sequential execution of ROMC. We observe
      that the parallel version runs almost 5 times faster.}
      \label{fig:exec_parallel}
\end{figure}
